Gene Expression Analysis
========================================================

Nitesh Turaga, Johns Hopkins University, Email: nturaga1@jhmi.edu 


Original Code by: **Luigi Marchionni, MD, PhD**, Johns Hopkins University, Email: marchion@jhu.edu



Class Content:
-------------------



**Class Date:  May 6th**

1. Affymetrix Gene expression data analysis

   
**Class Date: May 9th**

1. Gene Set Analysis
2. RNAseq Analysis 



Class WebPage:
-------------

[https://github.com/nitesh1989/GeneExpressionDataAnalysis.git](https://github.com/nitesh1989/GeneExpressionDataAnalysis.git)


Class Structure:
----------------

This is going to be a hands on training workshop, and we expect the students to follow along as we go through the code.
If anyone is stuck at any point or is unable to run some part of the code, we encourage you to ask us. This is important
to move forward in the workshop.


Instructions:
-------------

1. Please go to the link [https://github.com/nitesh1989/GeneExpressionDataAnalysis.git](https://github.com/nitesh1989/GeneExpressionDataAnalysis.git).

2. On the left bottom of your page, you have the link to **Download ZIP** . This will let you download the entire repository.

3. Unzip the file after downloading it. 

4. In each of the files, prior to running the R Code, you have to change the working directory. This is crucial.

5. Please install the following packages once you have R(version 3.0 and above) and RStudio running. You can copy and paste the code below to install and update the packages.

```{r installPackages,eval=FALSE}

installedPckgs <- installed.packages()[,"Package"]

### Here we define the list of desired libraries
pckgListBIOC <- c("affy", "limma", "hgu95a.db", "annotate","multtest","KEGG.db","GO.db","RTopper")

### We source the biocLite.R script from the Bioconductor website
source("http://bioconductor.org/biocLite.R")

### Load the packages, or install them from Bioconductor if needed
for (pckg in pckgListBIOC) {
    if (! pckg %in% installedPckgs) biocLite(pckg)
	require(pckg, character.only=TRUE)

}

```

I am printing my session information to show my R configuration when I ran this code.

```{r}
sessionInfo()
```


